---
title: "AfSIS Soil Spectra Preprocessing"
author: "Mario Vivero"
date: "12/10/2014"
output: html_document
---
AfSIS Soil Spectra Preprocessing
========================================================
This article is meant as a guide to spectra preprocessing based on two introductory books: Chemometrics - A Practical Guide by Beebe at. al. and Chemometrics with R by Wehrens using the data from the Africa Soil Property Prediction Challenge which can be downloaded on: http://www.kaggle.com/c/afsis-soil-properties/data


According to Wikipedia, signal preprocessing in chemometrics is also a critical component of almost all chemometric applications, particularly the use of signal pretreatments to condition data prior to calibration or classification. The techniques employed commonly in chemometrics are often closely related to those used in related fields[1]. Meaning that many of the techniques applied here can be also be applied to other areas. In my personal experience I've encountered spectral analysis in two areas: Neuroscience and Finance (Stock Markets) for which I am aware of the importance of signal preprocessing because an algorithm will output better results when the data does not contain noise and is corrected, even nowadays with advanced algorithms such as deep neural networks.

Deep neural networks (H20 implementation - http://h2o.ai/) were tested with and without spectra preprocessing. Deep neural nets with spectra preprocessing, even a simple method were superior than the results without preprocessing. Said process will not be shown here since that is not the goal of this article. For a pure "Deep Learning" solution please refer to the section about deep autoencoders found on:
http://deeplearning.net/tutorial/dA.html#daa
http://ufldl.stanford.edu/wiki/index.php/UFLDL_Tutorial

The first step is to load the libraries and data into R


Libraries, Functions and Data
------------

```{r results='hide', message=FALSE}
#Load/install required libraries
require('ggplot2')
require('reshape2')
require('stringr')
require('grid')
require('prospectr')
require('pls')
require('ptw')

#Load Data
train <- read.csv('training.csv', header = TRUE, stringsAsFactors = FALSE)
test <- read.csv('sorted_test.csv', header = TRUE, stringsAsFactors = FALSE)
#Data Transformation
train <- transform(train, Depth = as.numeric(as.factor(train$Depth)) - 1)
test <- transform(test, Depth = as.numeric(as.factor(test$Depth)) - 1)

```
In order to display grid plots using ggplot2 we are going to need two additional functions found on:
http://www.kaggle.com/c/afsis-soil-properties/forums/t/10184/first-derivative
and
http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_%28ggplot2%29/
after this, we will be ready to start exploring the effects of preprocessing in our data.

```{r results='hide'}
#Additional Functions
#Plotting Spectra
plotSpectra <- function(numberOfSamples, spectralData, subsample, dataDF){
  #based on http://www.kaggle.com/c/afsis-soil-properties/forums/t/10184/first-derivative
  ixs <- sample(which(1:nrow(dataDF) %in% subsample), numberOfSamples)
  trainRawSub <- melt(dataDF[ixs, ], id.vars = "PIDN", measure.vars = spectralData)
  trainRawSub$variable <- as.numeric(str_replace_all(trainRawSub$variable,"m",""))
  ggplot(trainRawSub, aes(x = variable, y = value, colour = PIDN)) + geom_line()
}

# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
#
#gotten from : http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_%28ggplot2%29/

multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  require(grid)  
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)
  numPlots = length(plots)
  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
  }
  if (numPlots==1) {
    print(plots[[1]])
  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))  
    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))      
      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```
Smoothing
------------
Smoothing is an important step because it attempts to remove noise. Noise is usually assumed to be high frequency data injected to the signal. First we will start visualizing the raw data for comparison.

```{r fig.width=16, fig.height=16}
#Signal Processing
#Spectra / CO2 and others
CO2SignalOR <- seq(which(names(train) == 'm2379.76'), which(names(train) == 'm2352.76'))
allSpectralDataOR <- seq(which(names(train) == 'm7497.96'), which(names(train) == 'm599.76'))
allSpectralDataNoCO2OR <- c(seq(which(names(train) == 'm7497.96'), which(names(train) == 'm2379.76')), 
                            seq(which(names(train) == 'm2352.76'), which(names(train) == 'm599.76')))
spatialPredictorsOR <- seq(which(names(train) == 'BSAN'), which(names(train) == 'TMFI'))
depthIxOR <- which(names(train) == 'Depth')
#Subsoil all spectra
allSpectraSubsoil <- plotSpectra(10, spectralData = allSpectralDataOR, subsample = which(train$Depth == 0), train)
#Topsoil all spectra
allSpectraTopsoil <- plotSpectra(10, spectralData = allSpectralDataOR, subsample = which(train$Depth == 1), train)
#Subsoil no CO2
SpectralDataNoCO2Subsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2OR, subsample = which(train$Depth == 0), train)
#Topsoil no CO2
SpectralDataNoCO2Topsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2OR, subsample = which(train$Depth == 1), train)

multiplot(allSpectraSubsoil, allSpectraTopsoil, SpectralDataNoCO2Subsoil, SpectralDataNoCO2Topsoil, cols = 2)
```
The first and most obvious smoothing transformation that also can be used as dimensional reduction is Binning. Averaging a certain amount of neighboring frequencies results also in smoothed data. This is the first method that uses a certain window to perform its average, the wider the window the smoother the result. If a window is to big then some features that weren't originally noise can dissapear. Here I will show binning with a window of 11 and with a window of 50. The binning function can be found in the *prospectr* package, as "binning".

```{r fig.width=16, fig.height=8}
#Binning
#w/ bins of window 11
trainBin11 <- binning(train[ , allSpectralDataOR], bin.size=10)
trainBin11 <- cbind(train[ , 'PIDN'], trainBin11, train[ , spatialPredictorsOR], train$Depth)
names(trainBin11)[1] <- 'PIDN'
names(trainBin11)[length(trainBin11)] <- 'Depth'
#Spectra / CO2 and others
allSpectralData <- seq(2, length(trainBin11) - 16)
spatialPredictors <- seq(which(names(trainBin11) == 'BSAN'), which(names(trainBin11) == 'TMFI'))
depthIx <- which(names(trainBin11) == 'Depth')#
#Subsoil all spectra
bin11SmootherallSpectraSubsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainBin11$Depth == 0), trainBin11)
#Topsoil all spectra
bin11SmootherallSpectraTopsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainBin11$Depth == 1), trainBin11)

multiplot(bin11SmootherallSpectraSubsoil, bin11SmootherallSpectraTopsoil, cols = 2)
```

```{r fig.width=16, fig.height=8}

#w/ bins of window 50
trainBin50 <- binning(train[ , allSpectralDataOR], bin.size=50)
trainBin50 <- cbind(train[ , 'PIDN'], trainBin50, train[ , spatialPredictorsOR], train$Depth)
names(trainBin50)[1] <- 'PIDN'
names(trainBin50)[length(trainBin50)] <- 'Depth'
#Spectra / CO2 and others
allSpectralData <- seq(2, length(trainBin50) - 16)
spatialPredictors <- seq(which(names(trainBin50) == 'BSAN'), which(names(trainBin50) == 'TMFI'))
depthIx <- which(names(trainBin50) == 'Depth')
#Subsoil all spectra
bin50SmootherallSpectraSubsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainBin50$Depth == 0), trainBin50)
#Topsoil all spectra
bin50SmootherallSpectraTopsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainBin50$Depth == 1), trainBin50)

multiplot(bin50SmootherallSpectraSubsoil, bin50SmootherallSpectraSubsoil, cols = 2)
```

Next I will implement the Running mean smoother using the running mean smoothing. This will apply an algorithm very similar to k nearest neighbor. The running mean function can be found in the *prospectr* package, as "movav".
```{r fig.width=16, fig.height=16}

#Running Mean Smoother
trainMeanSmoother <- as.data.frame(movav(train[ , allSpectralDataOR], w = 11))
trainMeanSmoother <- cbind(train[ , 'PIDN'], trainMeanSmoother, train[ , spatialPredictorsOR], train$Depth)
names(trainMeanSmoother)[1] <- 'PIDN'
names(trainMeanSmoother)[length(trainMeanSmoother)] <- 'Depth'

#Spectra / CO2 and others
CO2Signal <- seq(which(names(trainMeanSmoother) == 'm2379.76'), which(names(trainMeanSmoother) == 'm2352.76'))
allSpectralData <- seq(2, length(trainMeanSmoother) - 16)
allSpectralDataNoCO2 <- c(seq(2, which(names(trainMeanSmoother) == 'm2379.76')), 
                          seq(which(names(trainMeanSmoother) == 'm2352.76'), length(trainMeanSmoother) - 16))
spatialPredictors <- seq(which(names(trainMeanSmoother) == 'BSAN'), which(names(trainMeanSmoother) == 'TMFI'))
depthIx <- which(names(trainMeanSmoother) == 'Depth')

#Subsoil all spectra
meanSmootherallSpectraSubsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainMeanSmoother$Depth == 0), trainMeanSmoother)
#Topsoil all spectra
meanSmootherallSpectraTopsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainMeanSmoother$Depth == 1), trainMeanSmoother)
#Subsoil no CO2
meanSmootherSpectralDataNoCO2Subsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainMeanSmoother$Depth == 0), trainMeanSmoother)
#Topsoil no CO2
meanSmootherSpectralDataNoCO2Topsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainMeanSmoother$Depth == 1), trainMeanSmoother)

multiplot(meanSmootherallSpectraSubsoil, meanSmootherallSpectraTopsoil,
          meanSmootherSpectralDataNoCO2Subsoil, meanSmootherSpectralDataNoCO2Topsoil, cols = 2)
```
The most popular polynomial smoother is implemented as part of the Savitzky-Golay filter. The Savitzky-Golay filter can also implement derivatives, since we are only interested in seeing the effects of smoothing on the spectra we will left the parameter "m" which declares the derivative to 0.

```{r fig.width=16, fig.height=16}

#Running Polinomial Smoother (Savitzky-Golay filtering)
trainSGSmoother <- as.data.frame(savitzkyGolay(train[ , allSpectralDataOR], p = 3, w = 11, m = 0))
trainSGSmoother <- cbind(train[ , 'PIDN'], trainSGSmoother, train[ , spatialPredictorsOR], train$Depth)
names(trainSGSmoother)[1] <- 'PIDN'
names(trainSGSmoother)[length(trainSGSmoother)] <- 'Depth'

#Spectra / CO2 and others
CO2Signal <- seq(which(names(trainSGSmoother) == 'm2379.76'), which(names(trainSGSmoother) == 'm2352.76'))
allSpectralData <- seq(2, length(trainSGSmoother) - 16)
allSpectralDataNoCO2 <- c(seq(2, which(names(trainSGSmoother) == 'm2379.76')), 
                          seq(which(names(trainSGSmoother) == 'm2352.76'), length(trainSGSmoother) - 16))
spatialPredictors <- seq(which(names(trainSGSmoother) == 'BSAN'), which(names(trainSGSmoother) == 'TMFI'))
depthIx <- which(names(trainSGSmoother) == 'Depth')

#Subsoil all spectra
SGSmootherallSpectraSubsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainSGSmoother$Depth == 0), trainSGSmoother)
#Topsoil all spectra
SGSmootherallSpectraTopsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainSGSmoother$Depth == 1), trainSGSmoother)
#Subsoil no CO2
SGSmootherSpectralDataNoCO2Subsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainSGSmoother$Depth == 0), trainSGSmoother)
#Topsoil no CO2
SGSmootherSpectralDataNoCO2Topsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainSGSmoother$Depth == 1), trainSGSmoother)

multiplot(SGSmootherallSpectraSubsoil, SGSmootherallSpectraTopsoil,
          SGSmootherSpectralDataNoCO2Subsoil, SGSmootherSpectralDataNoCO2Topsoil, cols = 2)
```
Another option is the use of Fourier Filtering. More details on that can be found in  Chemometrics - A Practical Guide by Beebe at. al. P. 34.

Baseline Correction
------------
There are several approaches to Baseline Correction, here I will only include three. For explicit derivative calculations and running differences refer to Chemometrics - A Practical Guide by Beebe at. al. P. 36 - 45.

The recommended implementation by many is the Savitzky - Golay and Gorry with first, second or third derivative. The Savitzky - Golay filter with a derivative can be found in the *prospectr* package, as "savitzkyGolay" with parameter "m" > 0 (m determines the derivative).

```{r fig.width=16, fig.height=16}

#Baseline Correction
#Savitzky - Golay and Gorry with second dderivative
trainSGBasslineCor <- as.data.frame(savitzkyGolay(train[ , allSpectralDataOR], p = 3, w = 11, m = 2))
trainSGBasslineCor <- cbind(train[ , 'PIDN'], trainSGBasslineCor, train[ , spatialPredictorsOR], train$Depth)
names(trainSGBasslineCor)[1] <- 'PIDN'
names(trainSGBasslineCor)[length(trainSGBasslineCor)] <- 'Depth'

#Spectra / CO2 and others
CO2Signal <- seq(which(names(trainSGBasslineCor) == 'm2379.76'), which(names(trainSGBasslineCor) == 'm2352.76'))
allSpectralData <- seq(2, length(trainSGBasslineCor) - 16)
allSpectralDataNoCO2 <- c(seq(2, which(names(trainSGBasslineCor) == 'm2379.76')), 
                          seq(which(names(trainSGBasslineCor) == 'm2352.76'), length(trainSGBasslineCor) - 16))
spatialPredictors <- seq(which(names(trainSGBasslineCor) == 'BSAN'), which(names(trainSGBasslineCor) == 'TMFI'))
depthIx <- which(names(trainSGBasslineCor) == 'Depth')

#Subsoil all spectra
SGBassallSpectraSubsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainSGBasslineCor$Depth == 0), trainSGBasslineCor)
#Topsoil all spectra
SGBassallSpectraTopsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainSGBasslineCor$Depth == 1), trainSGBasslineCor)
#Subsoil no CO2
SGBassSpectralDataNoCO2Subsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainSGBasslineCor$Depth == 0), trainSGBasslineCor)
#Topsoil no CO2
SGBassSpectralDataNoCO2Topsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainSGBasslineCor$Depth == 1), trainSGBasslineCor)

multiplot(SGBassallSpectraSubsoil, SGBassallSpectraTopsoil,
          SGBassSpectralDataNoCO2Subsoil, SGBassSpectralDataNoCO2Topsoil, cols = 2)
```

The Multiplicative Scatter Correction is a preprocessing tool developed to correct the significant light-scattering problems in refrectance spectroscopy. Since then it has been found to be of more general use. Chemometrics - A Practical Guide by Beebe at. al. P. 46. the "msc" function can be found in the *pls* package. 

```{r fig.width=16, fig.height=16}

#MSC (Multiplicative Scatter Correction)
trainMSC <- as.data.frame(msc(as.matrix(train[ , allSpectralDataOR])))
trainMSC <- cbind(train[ , 'PIDN'], trainMSC, train[ , spatialPredictorsOR], train$Depth)
names(trainMSC)[1] <- 'PIDN'
names(trainMSC)[length(trainMSC)] <- 'Depth'

#Spectra / CO2 and others
CO2Signal <- seq(which(names(trainMSC) == 'm2379.76'), which(names(trainMSC) == 'm2352.76'))
allSpectralData <- seq(2, length(trainMSC) - 16)
allSpectralDataNoCO2 <- c(seq(2, which(names(trainMSC) == 'm2379.76')), 
                          seq(which(names(trainMSC) == 'm2352.76'), length(trainMSC) - 16))
spatialPredictors <- seq(which(names(trainMSC) == 'BSAN'), which(names(trainMSC) == 'TMFI'))
depthIx <- which(names(trainMSC) == 'Depth')

#Subsoil all spectra
MSCSpectraSubsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainMSC$Depth == 0), trainMSC)
#Topsoil all spectra
MSCSpectraTopsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainMSC$Depth == 1), trainMSC)
#Subsoil no CO2
MSCSpectraNoCO2Subsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainMSC$Depth == 0), trainMSC)
#Topsoil no CO2
MSCSpectraNoCO2Topsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainMSC$Depth == 1), trainMSC)

multiplot(MSCSpectraSubsoil, MSCSpectraTopsoil, MSCSpectraNoCO2Subsoil, MSCSpectraNoCO2Topsoil, cols = 2)
```

Another alternative is to use asymmetric least squares, where deviations above the fitted curve are not taken into account (or only with a very small weight). This is implemented in function baseline.corr in the *ptw* package, which returns a baseline-corrected signal. Internally, it uses the function "asysm"" to estimate the baseline. Chemometrics with R by Wehrens P. 20

```{r fig.width=16, fig.height=16}

#Asymmetric Least Squares
trainALS <- as.data.frame(baseline.corr(train[ , allSpectralDataOR]))
trainALS <- cbind(train[ , 'PIDN'], trainALS, train[ , spatialPredictorsOR], train$Depth)
names(trainALS)[1] <- 'PIDN'
names(trainALS)[length(trainALS)] <- 'Depth'

#Spectra / CO2 and others
CO2Signal <- seq(which(names(trainALS) == 'm2379.76'), which(names(trainALS) == 'm2352.76'))
allSpectralData <- seq(2, length(trainALS) - 16)
allSpectralDataNoCO2 <- c(seq(2, which(names(trainALS) == 'm2379.76')), 
                          seq(which(names(trainALS) == 'm2352.76'), length(trainALS) - 16))
spatialPredictors <- seq(which(names(trainALS) == 'BSAN'), which(names(trainALS) == 'TMFI'))
depthIx <- which(names(trainALS) == 'Depth')

#Subsoil all spectra
ALSSpectraSubsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainALS$Depth == 0), trainALS)
#Topsoil all spectra
ALSSpectraTopsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainALS$Depth == 1), trainALS)
#Subsoil no CO2
ALSDataNoCO2Subsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainALS$Depth == 0), trainALS)
#Topsoil no CO2
ALSDataNoCO2Topsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainALS$Depth == 1), trainALS)

multiplot(ALSSpectraSubsoil, ALSSpectraTopsoil, ALSDataNoCO2Subsoil, ALSDataNoCO2Topsoil, cols = 2)
```

Conclusion
------------

As it can be here seen, picking a certain smoothing procedure plus a baseline correction may be one solution. Another solution can be to use a Savitzky - Golay filter with a derivative in case the baseline doesn't have a very complex shape. The best way to put this is how Wehrens explained it: "Careful data pretreatment is essential – baselines may severely influence the results and should be removed before alignment... Often, the biggest gain in the alignment optimization is achieved by getting the prominent features in the right location. Sometimes, this dominance leads to suboptimal alignments. Also differences in intensity between sample and reference signals can distort the results."
As it was pointed out earlier, this write-up is only meant as a practical introduction to spectra preprocessing. To further your knowledge about this topic I highly recommend the books in the bibliography where you can find more advanced topics in preprocessing such as dynamic time warping and many others as well as algorithms and calibration tools.

Bibliography
------------
[1]https://en.wikipedia.org/wiki/Chemometrics#Other_techniques
[2]Beebe at. al., Chemometrics - A Practical Guide, March 1998, ISBN: 978-0-471-12451-1
[3]Wehrens R., Chemometrics with R Multivariate Data Analysis in the Natural Sciences and Life Sciences; 2011; ISBN 978-3-642-17840-5; e-ISBN 978-3-642-17841-2
