AfSIS Soil Spectra Preprocessing
========================================================
This article is meant as a guide to spectra preprocessing based on two introductory books: Chemometrics - A Prectical Guide by Beebe at. al. and Chemometrics with R by Wehrens using the data from the Africa Soil Property Prediction Challenge which can be downloaded on: http://www.kaggle.com/c/afsis-soil-properties/data


According to Wikipedia, signal preprocessing in chemometrics "is also a critical component of almost all chemometric applications, particularly the use of signal pretreatments to condition data prior to calibration or classification. The techniques employed commonly in chemometrics are often closely related to those used in related fields"[1]. Meaning that many of the techniques applied here can be also be applied to other areas. In my personal experience I've encountered spectral analysis in two areas: Neuroscience and Finance (Stock Markets) for which I am aware of the importance of signal preprocessing because an algorithm will output better results when the data does not contain noise and is corrected, even nowadays with advanced algorithms such as deep neural networks.

Deep neural networks (H20 implementation - http://h2o.ai/) were tested with and without spectra preprocessing. Deep neural nets with spectra preprocessing, even a simple method were superior than the results without preprocessing. Said process will not be shown here since that is not the goal of this article. For a pure "Deep Learning" solution please refer to the section about deep autoencoders found on:
http://deeplearning.net/tutorial/dA.html#daa
http://ufldl.stanford.edu/wiki/index.php/UFLDL_Tutorial

The first step is to load the libraries and data into R


Libraries, Functions and Data
------------

```{r results='hide', message=FALSE}
#Load/install required libraries
require('ggplot2')
require('reshape2')
require('stringr')
require('grid')
require('prospectr')
require('pls')
require('ptw')

#Load Data
train <- read.csv('training.csv', header = TRUE, stringsAsFactors = FALSE)
test <- read.csv('sorted_test.csv', header = TRUE, stringsAsFactors = FALSE)
#Data Transformation
train <- transform(train, Depth = as.numeric(as.factor(train$Depth)) - 1)
test <- transform(test, Depth = as.numeric(as.factor(test$Depth)) - 1)

```
In order to display grid plots using ggplot2 we are going to need two additional functions found on:
*http://www.kaggle.com/c/afsis-soil-properties/forums/t/10184/first-derivative*
and
*http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_%28ggplot2%29/*
after this, we will be ready to start exploring the effects of preprocessing in our data.

```{r results='hide'}
#Additional Functions
#Plotting Spectra
plotSpectra <- function(numberOfSamples, spectralData, subsample, dataDF){
  #based on http://www.kaggle.com/c/afsis-soil-properties/forums/t/10184/first-derivative
  ixs <- sample(which(1:nrow(dataDF) %in% subsample), numberOfSamples)
  trainRawSub <- melt(dataDF[ixs, ], id.vars = "PIDN", measure.vars = spectralData)
  trainRawSub$variable <- as.numeric(str_replace_all(trainRawSub$variable,"m",""))
  ggplot(trainRawSub, aes(x = variable, y = value, colour = PIDN)) + geom_line()
}

# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
#
#gotten from : http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_%28ggplot2%29/

multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  require(grid)  
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)
  numPlots = length(plots)
  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
  }
  if (numPlots==1) {
    print(plots[[1]])
  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))  
    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))      
      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```
Smoothing
------------
Smoothing is an important step because it attempts to remove noise. Noise is usually assumed to be high frecuency data injected to the signal. 

#Transform it to H2O compatible file
#Make a Shuffled CSV to train h2o models
set.seed(1010)
randomSubset <- sample.int(nrow(train), nrow(train)) #full data
trainShuffled <- train[randomSubset, ]
trainShuffled$P <- log(trainShuffled$P + 2)
write.csv(trainShuffled, file = paste0(dataDirectory, 'trainingShuffled.csv'), row.names = FALSE)

#Test Data
write.csv(test, file = paste0(dataDirectory, 'testNumeric.csv'), row.names = FALSE)

#Signal Processing
#Spectra / CO2 and others
CO2SignalOR <- seq(which(names(train) == 'm2379.76'), which(names(train) == 'm2352.76'))
allSpectralDataOR <- seq(which(names(train) == 'm7497.96'), which(names(train) == 'm599.76'))
allSpectralDataNoCO2OR <- c(seq(which(names(train) == 'm7497.96'), which(names(train) == 'm2379.76')), 
                            seq(which(names(train) == 'm2352.76'), which(names(train) == 'm599.76')))
spatialPredictorsOR <- seq(which(names(train) == 'BSAN'), which(names(train) == 'TMFI'))
depthIxOR <- which(names(train) == 'Depth')



#Subsoil all spectra
allSpectraSubsoil <- plotSpectra(10, spectralData = allSpectralDataOR, subsample = which(train$Depth == 0), train)
#Topsoil all spectra
allSpectraTopsoil <- plotSpectra(10, spectralData = allSpectralDataOR, subsample = which(train$Depth == 1), train)
#Subsoil no CO2
SpectralDataNoCO2Subsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2OR, subsample = which(train$Depth == 0), train)
#Topsoil no CO2
SpectralDataNoCO2Topsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2OR, subsample = which(train$Depth == 1), train)

multiplot(allSpectraSubsoil, allSpectraTopsoil, SpectralDataNoCO2Subsoil, SpectralDataNoCO2Topsoil, cols = 2)

#--------------------------------
#Transformations
#Smoothing
#Binning
#w/ 11 bins
trainBin11 <- binning(train[ , allSpectralDataOR], bin.size=10)
trainBin11 <- cbind(train[ , 'PIDN'], trainBin11, train[ , spatialPredictorsOR], train$Depth)
names(trainBin11)[1] <- 'PIDN'
names(trainBin11)[length(trainBin11)] <- 'Depth'
#Spectra / CO2 and others
allSpectralData <- seq(2, length(trainBin11) - 16)
spatialPredictors <- seq(which(names(trainBin11) == 'BSAN'), which(names(trainBin11) == 'TMFI'))
depthIx <- which(names(trainBin11) == 'Depth')#
#Subsoil all spectra
bin11SmootherallSpectraSubsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainBin11$Depth == 0), trainBin11)
#Topsoil all spectra
bin11SmootherallSpectraTopsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainBin11$Depth == 1), trainBin11)

multiplot(bin11SmootherallSpectraSubsoil, bin11SmootherallSpectraTopsoil, cols = 2)

#w/ 50 bins
trainBin50 <- binning(train[ , allSpectralDataOR], bin.size=50)
trainBin50 <- cbind(train[ , 'PIDN'], trainBin50, train[ , spatialPredictorsOR], train$Depth)
names(trainBin50)[1] <- 'PIDN'
names(trainBin50)[length(trainBin50)] <- 'Depth'
#Spectra / CO2 and others
allSpectralData <- seq(2, length(trainBin50) - 16)
spatialPredictors <- seq(which(names(trainBin50) == 'BSAN'), which(names(trainBin50) == 'TMFI'))
depthIx <- which(names(trainBin50) == 'Depth')
#Subsoil all spectra
bin50SmootherallSpectraSubsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainBin50$Depth == 0), trainBin50)
#Topsoil all spectra
bin50SmootherallSpectraTopsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainBin50$Depth == 1), trainBin50)

multiplot(bin50SmootherallSpectraSubsoil, bin50SmootherallSpectraSubsoil, cols = 2)

#Running Mean Smoother
trainMeanSmoother <- as.data.frame(movav(train[ , allSpectralDataOR], w = 11))
trainMeanSmoother <- cbind(train[ , 'PIDN'], trainMeanSmoother, train[ , spatialPredictorsOR], train$Depth)
names(trainMeanSmoother)[1] <- 'PIDN'
names(trainMeanSmoother)[length(trainMeanSmoother)] <- 'Depth'

#Spectra / CO2 and others
CO2Signal <- seq(which(names(trainMeanSmoother) == 'm2379.76'), which(names(trainMeanSmoother) == 'm2352.76'))
allSpectralData <- seq(2, length(trainMeanSmoother) - 16)
allSpectralDataNoCO2 <- c(seq(2, which(names(trainMeanSmoother) == 'm2379.76')), 
                          seq(which(names(trainMeanSmoother) == 'm2352.76'), length(trainMeanSmoother) - 16))
spatialPredictors <- seq(which(names(trainMeanSmoother) == 'BSAN'), which(names(trainMeanSmoother) == 'TMFI'))
depthIx <- which(names(trainMeanSmoother) == 'Depth')

#Subsoil all spectra
meanSmootherallSpectraSubsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainMeanSmoother$Depth == 0), trainMeanSmoother)
#Topsoil all spectra
meanSmootherallSpectraTopsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainMeanSmoother$Depth == 1), trainMeanSmoother)
#Subsoil no CO2
meanSmootherSpectralDataNoCO2Subsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainMeanSmoother$Depth == 0), trainMeanSmoother)
#Topsoil no CO2
meanSmootherSpectralDataNoCO2Topsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainMeanSmoother$Depth == 1), trainMeanSmoother)

multiplot(meanSmootherallSpectraSubsoil, meanSmootherallSpectraTopsoil,
          meanSmootherSpectralDataNoCO2Subsoil, meanSmootherSpectralDataNoCO2Topsoil, cols = 2)

#Running Polinomial Smoother (Savitzky-Golay filtering)
trainSGSmoother <- as.data.frame(savitzkyGolay(train[ , allSpectralDataOR], p = 3, w = 11, m = 0))
trainSGSmoother <- cbind(train[ , 'PIDN'], trainSGSmoother, train[ , spatialPredictorsOR], train$Depth)
names(trainSGSmoother)[1] <- 'PIDN'
names(trainSGSmoother)[length(trainSGSmoother)] <- 'Depth'

#Spectra / CO2 and others
CO2Signal <- seq(which(names(trainSGSmoother) == 'm2379.76'), which(names(trainSGSmoother) == 'm2352.76'))
allSpectralData <- seq(2, length(trainSGSmoother) - 16)
allSpectralDataNoCO2 <- c(seq(2, which(names(trainSGSmoother) == 'm2379.76')), 
                          seq(which(names(trainSGSmoother) == 'm2352.76'), length(trainSGSmoother) - 16))
spatialPredictors <- seq(which(names(trainSGSmoother) == 'BSAN'), which(names(trainSGSmoother) == 'TMFI'))
depthIx <- which(names(trainSGSmoother) == 'Depth')

#Subsoil all spectra
SGSmootherallSpectraSubsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainSGSmoother$Depth == 0), trainSGSmoother)
#Topsoil all spectra
SGSmootherallSpectraTopsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainSGSmoother$Depth == 1), trainSGSmoother)
#Subsoil no CO2
SGSmootherSpectralDataNoCO2Subsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainSGSmoother$Depth == 0), trainSGSmoother)
#Topsoil no CO2
SGSmootherSpectralDataNoCO2Topsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainSGSmoother$Depth == 1), trainSGSmoother)

multiplot(SGSmootherallSpectraSubsoil, SGSmootherallSpectraTopsoil,
          SGSmootherSpectralDataNoCO2Subsoil, SGSmootherSpectralDataNoCO2Topsoil, cols = 2)

#Baseline Correction
#Savitzky - Golay and Gorry with second dderivative
trainSGBasslineCor <- as.data.frame(savitzkyGolay(train[ , allSpectralDataOR], p = 3, w = 11, m = 2))
trainSGBasslineCor <- cbind(train[ , 'PIDN'], trainSGBasslineCor, train[ , spatialPredictorsOR], train$Depth)
names(trainSGBasslineCor)[1] <- 'PIDN'
names(trainSGBasslineCor)[length(trainSGBasslineCor)] <- 'Depth'

#Spectra / CO2 and others
CO2Signal <- seq(which(names(trainSGBasslineCor) == 'm2379.76'), which(names(trainSGBasslineCor) == 'm2352.76'))
allSpectralData <- seq(2, length(trainSGBasslineCor) - 16)
allSpectralDataNoCO2 <- c(seq(2, which(names(trainSGBasslineCor) == 'm2379.76')), 
                          seq(which(names(trainSGBasslineCor) == 'm2352.76'), length(trainSGBasslineCor) - 16))
spatialPredictors <- seq(which(names(trainSGBasslineCor) == 'BSAN'), which(names(trainSGBasslineCor) == 'TMFI'))
depthIx <- which(names(trainSGBasslineCor) == 'Depth')

#Subsoil all spectra
SGBassallSpectraSubsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainSGBasslineCor$Depth == 0), trainSGBasslineCor)
#Topsoil all spectra
SGBassallSpectraTopsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainSGBasslineCor$Depth == 1), trainSGBasslineCor)
#Subsoil no CO2
SGBassSpectralDataNoCO2Subsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainSGBasslineCor$Depth == 0), trainSGBasslineCor)
#Topsoil no CO2
SGBassSpectralDataNoCO2Topsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainSGBasslineCor$Depth == 1), trainSGBasslineCor)

multiplot(SGBassallSpectraSubsoil, SGBassallSpectraTopsoil,
          SGBassSpectralDataNoCO2Subsoil, SGBassSpectralDataNoCO2Topsoil, cols = 2)


#MSC (Multiplicative Scatter Correction)
trainMSC <- as.data.frame(msc(as.matrix(train[ , allSpectralDataOR])))
trainMSC <- cbind(train[ , 'PIDN'], trainMSC, train[ , spatialPredictorsOR], train$Depth)
names(trainMSC)[1] <- 'PIDN'
names(trainMSC)[length(trainMSC)] <- 'Depth'

#Spectra / CO2 and others
CO2Signal <- seq(which(names(trainMSC) == 'm2379.76'), which(names(trainMSC) == 'm2352.76'))
allSpectralData <- seq(2, length(trainMSC) - 16)
allSpectralDataNoCO2 <- c(seq(2, which(names(trainMSC) == 'm2379.76')), 
                          seq(which(names(trainMSC) == 'm2352.76'), length(trainMSC) - 16))
spatialPredictors <- seq(which(names(trainMSC) == 'BSAN'), which(names(trainMSC) == 'TMFI'))
depthIx <- which(names(trainMSC) == 'Depth')

#Subsoil all spectra
MSCSpectraSubsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainMSC$Depth == 0), trainMSC)
#Topsoil all spectra
MSCSpectraTopsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainMSC$Depth == 1), trainMSC)
#Subsoil no CO2
MSCSpectraNoCO2Subsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainMSC$Depth == 0), trainMSC)
#Topsoil no CO2
MSCSpectraNoCO2Topsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainMSC$Depth == 1), trainMSC)

multiplot(MSCSpectraSubsoil, MSCSpectraTopsoil, MSCSpectraNoCO2Subsoil, MSCSpectraNoCO2Topsoil, cols = 2)

#Asymmetric Least Squares
trainALS <- as.data.frame(baseline.corr(train[ , allSpectralDataOR]))
trainALS <- cbind(train[ , 'PIDN'], trainALS, train[ , spatialPredictorsOR], train$Depth)
names(trainALS)[1] <- 'PIDN'
names(trainALS)[length(trainALS)] <- 'Depth'

#Spectra / CO2 and others
CO2Signal <- seq(which(names(trainALS) == 'm2379.76'), which(names(trainALS) == 'm2352.76'))
allSpectralData <- seq(2, length(trainALS) - 16)
allSpectralDataNoCO2 <- c(seq(2, which(names(trainALS) == 'm2379.76')), 
                          seq(which(names(trainALS) == 'm2352.76'), length(trainALS) - 16))
spatialPredictors <- seq(which(names(trainALS) == 'BSAN'), which(names(trainALS) == 'TMFI'))
depthIx <- which(names(trainALS) == 'Depth')

#Subsoil all spectra
ALSSpectraSubsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainALS$Depth == 0), trainALS)
#Topsoil all spectra
ALSSpectraTopsoil <- plotSpectra(10, spectralData = allSpectralData, subsample = which(trainALS$Depth == 1), trainALS)
#Subsoil no CO2
ALSDataNoCO2Subsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainALS$Depth == 0), trainALS)
#Topsoil no CO2
ALSDataNoCO2Topsoil <- plotSpectra(10, spectralData = allSpectralDataNoCO2, subsample = which(trainALS$Depth == 1), trainALS)

multiplot(ALSSpectraSubsoil, ALSSpectraTopsoil, ALSDataNoCO2Subsoil, ALSDataNoCO2Topsoil, cols = 2)

Bibliography
------------
[1]https://en.wikipedia.org/wiki/Chemometrics#Other_techniques
[2]Beebe at. al., Chemometrics - A Prectical Guide, March 1998, ISBN: 978-0-471-12451-1
[3]Wehrens R., Chemometrics with R Multivariate Data Analysis in the Natural Sciences and Life Sciences; 2011; ISBN 978-3-642-17840-5; e-ISBN 978-3-642-17841-2
